---
title: "Damages of natural events to economy and human lives in US"
output: 
  html_document:
    keep_md: true
---

```{r echo=FALSE}
# Format numeric values appropriately
# see http://stackoverflow.com/questions/30888631/knitr-displaying-digits-of-an-integer-without-scientific-notation
inline_hook <- function(x){
  if(is.numeric(x)){
    paste(format(x))
  }
}
knitr::knit_hooks$set(inline=inline_hook)
```

## Synopsis
This study uses the data between January 1996 and November 2011 provided by the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database to address two questions:

1. Across the United States, which types of events are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?
 
To address the first question we look at the total number of injuries and fatalities caused by weather events by type. Among those we select the top 5 event types.

To address the second one we take a similar approach looking at the total economic loss (expressed as damages to properties and crops) and selecting the top five events that caused highest losses. 

Our analysis shows that :

1. considering fatalities and injuries, **Excessive heat, Tornados and Flood** can be considered the most harmful events to population health; 
2. **Flood, Hurricanes and Storm** as the worst extreme weather events in terms of economic losses.

## Data processing
This section describes data load and processing before its use.

### Required library
```{r message=FALSE, warning=FALSE}
require(data.table)
require(lubridate)
require(stringdist)
require(dplyr)
require(lattice)
require(gridExtra)
```

### Functions
#### `apply_exp(value, exp)`
Take a `value` and apply an exponent `exp`. Base of the exponent is 10.
```{r}
apply_exp <- function(value, exp) {
  switch(
    exp,
    H={
      return(value * 100);
    },
    K={
      return(value * 1000);
    },
    M={
      return(value * 1000000);
    },
    B={
      return(value * 1000000000);
    },
    "0"={
      return(value);
    },
    "1"={
      return(value * 10);
    },
    "2"={
      return(value * 10^2);
    },
    "3"={
      return(value * 10^3);
    },
    "4"={
      return(value * 10^4);
    },
    "5"={
      return(value * 10^5);
    },
    "6"={
      return(value * 10^6);
    },
    "7"={
      return(value * 10^7);
    },
    "8"={
      return(value * 10^7);
    },
    "9"={
      return(value * 10^7);
    },
    {
      # Junk data, do nothing
      return(value);
    }
  )
}
```

#### `print_perc(n)`
Print a number `n` as percentage.
```{r eval=TRUE}
print_perc <- function(n) {
  paste0(
    formatC(
      100 * n,
      format = "f",
      digits = 2
    ),
    "%"
  )
}
```

### Data load
Here we load the data storm database selecting only relevant columns:
```{r cache=TRUE}
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
destFile <- "repdata-data-StormData.csv.bz2"

# Download
if (!file.exists(destFile)) {
  res <- tryCatch(
    download.file(fileUrl, method = "libcurl", destfile = destFile),
    error = function(e)
      1
  )
}

colClasses = c(
  "NULL",
  "character", # BGN_DATE
  rep("NULL", 5),
  "character",  # EVTYPE
  rep("NULL", 14),
  rep("numeric", 3), # FATALITIES, INJURIES, PROPDMG
  "character", # PROPDMGEXP: Factor w/ 19 levels
  "numeric",   # CROPDMG
  "character", # CROPDMGEXP: Factor w/ 9 levels
  rep("NULL", 9)
)

# Read data - only relevant columns
df <- read.table(
  bzfile(destFile),
  sep = ",",
  header = TRUE,
  quote = "\"",
  colClasses = colClasses
)

colnames(df) <- c(
  "BGN_DATE",
  "EVTYPE",
  "FATALITIES",
  "INJURIES",
  "PROPDMG",
  "PROPDMGEXP",
  "CROPDMG",
  "CROPDMGEXP"
)

df$BGN_DATE <- as.Date(df$BGN_DATE, "%m/%d/%Y")
```

The events in the storm database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. In particular, only Tornados were recorded until January 1955 and until January 1996 only Tornados, Thunderstorms, Wind and Hails. That makes only 4 out of 48 official event types. Since we are comparing the effects of weather events and we have no history between 1950 and 1996 of 44 of them, we have decided to **use only records from January 1996 to November 2011**.

```{r}
df <- subset(df, BGN_DATE > as.Date("1996-01-01") )
```

### Data cleaning
Data in the storm database is not clean. To have it in better shape we make some transformations.

First, we convert all relevant fields to uppercase to avoid case sensitiviness issues. 
```{r}
df$EVTYPE     <- toupper(df$EVTYPE)
df$PROPDMGEXP <- toupper(df$PROPDMGEXP)
df$CROPDMGEXP <- toupper(df$CROPDMGEXP)
```

Fatalities and injuries are integers so we can parse them as such.
```{r}
df$FATALITIES <- as.integer(df$FATALITIES)
df$INJURIES   <- as.integer(df$INJURIES)
```

Field `PROPDMGEXP` and `CROPDMGEXP` record the exponent of the numeric value found in the field `PROPDMG` and `CROPDMG`. We want to have a plain number to ease computation and plotting. We do this using the following function and applying it to `PROPDMG` and `CROPDMG` columns:
```{r}
df$PROPDMG <- mapply(apply_exp, df$PROPDMG, df$PROPDMGEXP)
df$CROPDMG <- mapply(apply_exp, df$CROPDMG, df$CROPDMGEXP)
```

#### Cleaning of `EVTYPE`
The field `EVTYPE`, that records the type of extreme natural event, can assume values from a predefined list of event types. This list can be found at http://www.nws.noaa.gov/directives/sym/pd01016005curr.pdf and contains no 48 different event types. We would expect `EVTYPE` to contain only 48 unique values, actually it does contain 
`r length(unique(df$EVTYPE))`.Some clean up is therefore needed to match the highest possible number of correct event types against the official list. To achieve this we first load the official event types list that will be used to match `EVTYPE` records against.

```{r cache=TRUE}
# Official event types
evtypesFile <- "evtypes.csv"
evtypes     <- readLines(evtypesFile)
evtypes     <- toupper(evtypes)
```


Second, we perform a number data transformations. Here few substitutions to correct some records not matching any of the official event type:

```{r}
# Clean some EVTYPE
df$EVTYPE <- gsub("TSTM", "THUNDERSTORM", df$EVTYPE)
df$EVTYPE <- gsub(".*THUNDERSTORM WIND.*", "THUNDERSTORM WIND", df$EVTYPE)
df$EVTYPE <- gsub("URBAN/SML STREAM FLD", "FLOOD", df$EVTYPE)
df$EVTYPE <- gsub(".*FLOOD.*", "FLOOD", df$EVTYPE)
df$EVTYPE <- gsub("^FOG$", "DENSE FOG", df$EVTYPE)
df$EVTYPE <- gsub("^SNOW$", "HEAVY SNOW", df$EVTYPE)
df$EVTYPE <- gsub("^WIND$", "HIGH WIND", df$EVTYPE)
df$EVTYPE <- gsub(".*SURF.*", "HIGH SURF", df$EVTYPE)
```

here we use approximate match to automatically match records in `EVTYPE` against those in the official list. If a match occurs we update `ETYPE` with the event in the official list, if it does not we update it with `NA`:
```{r}
df$EVTYPE <- evtypes[amatch(df$EVTYPE, evtypes, method = "soundex")]
```

The percentage of total `EVTYPE` records matched against the official 48 long list after the aformentioned transformations is 

```{r}
print_perc(1 - sum(is.na(df$EVTYPE))/nrow(df))
```

## Results
In this section we present the results. We start computing the averages for `FATALITIES`, `INJURIES`, `PROPDMG` and `CROPDMG` aggregated by event types.

```{r cache=TRUE}
# Total damage summary
dmg_by_evtype <-
  df %>%
  group_by(EVTYPE) %>%
  summarise(
    total_fatalities = sum(FATALITIES),
    total_injuries   = sum(INJURIES),
    total_propdmg    = sum(PROPDMG),
    total_cropdmg    = sum(CROPDMG)
  )
```

We will use `FATALITIES` and `INJURIES` to explore which events are most harmful with respect to population health and `PROPDMG` and `CROPDMG` to understand the economic loss cause by each event.

### Most harmful types of events with respect to population health
Since we are interested in understanding whose events are the most harmful to population health we select only the top 5 events ordered by the average of fatalities and injuries caused.

```{r cache=TRUE}
# Plots for threats to human health
top_evtypes_by_fatalities <-
  head(
    arrange(
      dmg_by_evtype,
      desc(total_fatalities)
    ),
    5
  )

# This is required to enforce order on x axis since
# lattice reorders the labes alphabetically
x_labels <- 
  factor(
    top_evtypes_by_fatalities$EVTYPE, 
    levels = unique(top_evtypes_by_fatalities$EVTYPE)
  )

## Fatalities
fatalities_plot <- barchart(
  total_fatalities ~ x_labels,
  data=top_evtypes_by_fatalities,
  main = "Total fatalities by event type (top 5)",
  ylab = "Fatalities",
  scales=list(x=list(labels=x_labels, cex=0.65))
)

## Injuries
top_evtypes_by_injuries <- 
  head(
    arrange(dmg_by_evtype, desc(total_injuries))
    , 5
  )

x_labels <- 
  factor(
    top_evtypes_by_injuries$EVTYPE, 
    levels = unique(top_evtypes_by_injuries$EVTYPE)
  )

injuries_plot <- barchart(
  total_injuries ~ x_labels,
  data=top_evtypes_by_injuries,
  main = "Total injuries by event type (top 5)",
  ylab = "Injuries",
  scales=list(x=list(labels=x_labels, cex=0.65))
)

grid.arrange(fatalities_plot, injuries_plot, ncol=1)
```

The first plot (*Total fatalities by event type (top 5)*) shows that Excessive heat has caused the highest number of fatalities. The first three event types (Excessive Heat, Tornados and Floods) account for:
```{r cache=TRUE}
top_three_events_fatalities <- sum(top_evtypes_by_fatalities[1:3,2])
all_fatalities <- sum(dmg_by_evtype$total_fatalities)

print_perc(top_three_events_fatalities/all_fatalities)
```
of all fatalities.

The second plot (*Total injuries by event type (top 5)*) shows Tornados to be the first cause of injuries accounting for  
```{r eval=TRUE, cache=TRUE}
tornados_injuries <- sum(top_evtypes_by_injuries$total_injuries[1])
all_injuries <- sum(dmg_by_evtype$total_injuries)

print_perc(tornados_injuries/all_injuries)
```
of all injuries, while the first three event types account for 

```{r cache=TRUE}
top_three_events_injuries <- sum(top_evtypes_by_injuries[1:3,3])

print_perc(top_three_events_injuries/all_injuries)
```

of the total.

### Types of events that have the greatest economic consequences
To respond to the second question we are interested in understanding whose events caused the highest damages. Similarly to the previous case, we select only the top 5 events types ordered by the total damages to properties (`PROPDMG`) and to crops (`CROPDMG`).

```{r cache=TRUE}
## Plots for economic losses
top_evtypes_by_economic_dmg <-
  head(
    arrange(
      dmg_by_evtype,
      desc(total_propdmg+total_cropdmg)
    ),
    5
  )

x_labels <-
  factor(
    top_evtypes_by_economic_dmg$EVTYPE,
    levels = unique(top_evtypes_by_economic_dmg$EVTYPE)
  )

barchart(
  total_propdmg + total_cropdmg ~ x_labels,
  data=top_evtypes_by_economic_dmg,
  stack = TRUE,
  main = "Total damage by event type (top 5)",
  xlab = "Weather event type",
  ylab = "Damage",
  auto.key=list(
    corner = c(.98, .98),
    text = c("Property dmg", "Crop dmg")
  ),
  scales=list(x=list(labels=x_labels, rot=45, cex=0.65))
)
```